Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-08-03 paper.tex
typos fixed

In this paper, we will investigate in some more detail the potential
of several of the most promising galaxy clusters to produce an
annihilation gamma-ray yield which could be observable with present
and planned gamma-ray detectors. For convenience, we summarize in
Table~\ref{tab:flux_tab} our results for the gamma-ray flux from four
promising clusters with a high gamma-ray yield. We will in the
following derive the emission from two different DM models and a
hadronic CR model to show that Virgo has the highest gamma-ray flux
from annihilating DM. \C{However, if the boost from Sommerfeld
  enhancement (SFE) is small then the gamma-ray emission induced by
  CRs is dominating most clusters with the exception of a few
  clusters; most notably Fornax that shows the largest ratio of DM
  annihilation-to-CR flux. In particular, we identify Fornax and Virgo
  as prime targets for DM observations,} Perseus for CR induced
emission and we additionally decide in favor of the well studied
cluster Coma for comparison. While studying these four clusters in
great detail, we complement the analysis by calculating the expected
gamma-ray fluxes of a nearby X-ray flux complete sample of galaxy
clusters from the Rosat all-sky survey (extended HIghest X-ray FLUx
Galaxy Cluster Sample, HIFLUGCS,
\citealt{2002ApJ...567..716R,2007A&A...466..805C}).\footnote{Note that
  we have added the Virgo cluster to the sample that we nevertheless
  refer to as extended HIFLUGCS catalogue in the following.}  For
previous work related to DM in clusters, see, e.g.,
\cite{2006A&A...455...21C,2009PhRvD..80b3005J,2011arXiv1104.3530S,2011ApJ...726L...6C,2010JCAP...05..025A}
and for recent work on CRs in clusters see
\cite{2010MNRAS.409..449P,2011MNRAS.410..127B,2008MNRAS.385.1211P,2009JCAP...08..002K,2010MNRAS.407.1565D}.

@@

In supernovae remnants and on scales of galaxies, there are convincing
evidences of non-thermal populations. Especially, in the MW, the
cosmic rays are observed directly as well as indirectly through radio,
X-ray, and gamma-ray emission. On larger scales of the order of few
100~kpc up to Mpc, there are currently a vast number of observations
of radio emission coming from radio mini halos in the center of
cooling flow clusters, radio relics in the periphery of clusters
\cite{2004rcfg.proc..335K} as well as smooth giant radio halos that
have been observed in more than 50 clusters
\cite{2003ASPC..301..143F,2008SSRv..134...93F}. This type of emission
is expected in clusters since the formation process of a galaxy
cluster is a very energetic processes that induces both turbulence as
well as frequently occurring merging and accretion shocks which both
are thought to accelerate relativistic non-thermal protons and radio
emitting electrons to high energies. Thus most of the observed radio
emission in clusters is expected to emerge from cosmic-ray electron
induced radio synchrotron emission. The precise origin of these
electrons, in especially relics and giant radio halos, is still not
settled. One possible scenario for the production of these electrons
are hadronic CR interactions with ambient gas protons which results in
the production of both charged and neutral pions that decay into
electrons, neutrinos, and gamma-rays (see
\cite{1980ApJ...239L..93D,1982AJ.....87.1266V,1999APh....12..169B,2000A&A...362..151D,2001ApJ...559...59M,2003MNRAS.342.1009M,2004A&A...413...17P,2004MNRAS.352...76P,2008MNRAS.385.1211P,2008MNRAS.385.1242P,2009JCAP...09..024K,2010MNRAS.401...47D,2010MNRAS.407.1565D,2010ApJ...722..737K,2011A&A...527A..99E}
).\footnote{An alternative scenario is the second order turbulent
  re-acceleration through the interaction of a previously injected
  relativistic population of electrons by supernova driven winds or
  active galactic nuclei with plasma waves and
  magneto-turbulence. \protect
  \cite{1987A&A...182...21S,1993ApJ...406..399G,2004MNRAS.350.1174B,2005MNRAS.363.1173B,2007MNRAS.378..245B,2011MNRAS.410..127B,2009A&A...507..661B}}
Supporting evidence comes from the smoothness of the extended radio
emission that often resembles that observed in thermal X-rays. This
can be easily achieve in the hadronic model since the long cooling
time of cosmic ray protons (CRs) allows for an extended and large
population of CRs to build up over the cluster history
\citep{1996SSRv...75..279V, 1997ApJ...477..560E,
  1997ApJ...487..529B}. The production of these secondaries depend
both on the gas and CR densities in the cluster, where the CR
density roughly traces the gas outside the core regime and is slightly
enhanced in the center.  This density scaling implies that clusters
are great targets for Cherenkov telescopes with a high sensitivity for
the central parts of nearby clusters. Detecting the cluster gamma-ray
emission is crucial in this respect as it potentially provides the
unique and unambiguous evidence of CR populations in clusters through
observing the $\pi^0$ bump at about 100~MeV in the spectra.

@@

As expected, the analytical CR model accounts for the ``scatter'' to
the scaling relation and biases the gamma-ray flux high for prominent
cool core clusters by up to a factor of 10 relative to the expectation
from the scaling relation. This is due to the high target gas
densities and associated CR densities in those cool cores that shape
the spatial emission characteristic to be very similar to that
observed in thermal X-rays. We caution that these predictions only
apply for the case of negligible active CR transport such as CR
streaming and diffusion relative to the gas. Gamma-ray fluxes might be
considerably smaller in those cluster that do not show an extended
diffuse radio(-mini) halo that might point to a centrally concentrated
CR distribution \citep{2011A&A...527A..99E}. The CR spectrum $E^2
dN/dE$ shows the characteristic pion bump at energies around
1~GeV. \C{It is expected to dominate the DM annihilation signal of BM
  models for clusters/groups, although we find that the gamma-ray flux
  induced by the BM is higher for about 1/5 of the clusters in the
  HIFLUGCS catalogue, where Fornax and M49 have the highest signal to
  noise. We find that \Fermi-LAT will be able to start constraining
  these DM models within five years using clusters of galaxies.}
\del{It is expected to dominate the DM annihilation signal of BM
  models for all clusters/groups but Fornax.}

Combining extended \Fermi upper limits and our model flux predictions,
we limit the relative CR pressure, $X_\CR = P_\CR/P_\rmn{th}$, in 32
nearby bright galaxy clusters of the \Fermi sample. \C{The best limits
  are found in Norma and Coma of the order of 3\%, with typical limits
  around 10\%. This is comparable to those inferred by the \Fermi
  collaboration \citep{2010ApJ...717L..71A} and mainly due to the
  differently assumed Bayesian prior on the CR pressure which implies
  a different cluster mass dependence of the resulting limits on
  $X_\CR$.}

>>>>

In this paper, we will investigate in some more detail the potential
of several of the most promising galaxy clusters to produce an
annihilation gamma-ray yield which could be observable with present
and planned gamma-ray detectors. For convenience, we summarize in
Table~\ref{tab:flux_tab} our results for the gamma-ray flux from four
promising clusters with a high gamma-ray yield. We will in the
following derive the emission from two different DM models and a
hadronic CR model to show that Virgo has the highest gamma-ray flux
from annihilating DM. \C{However, if the boost from Sommerfeld
  enhancement (SFE) is small then the gamma-ray emission induced by
  CRs is dominating most clusters with the exception of a few
  clusters; most notably Fornax that shows the largest ratio of DM
  annihilation-to-CR flux. In particular, we identify Fornax and Virgo
  as prime targets for DM observations,} Perseus for CR induced
emission and we additionally decide in favor of the well studied
cluster Coma for comparison. While studying these four clusters in
great detail, we complement the analysis by calculating the expected
gamma-ray fluxes of a nearby X-ray flux complete sample of galaxy
clusters from the Rosat all-sky survey (extended HIghest X-ray FLUx
Galaxy Cluster Sample, HIFLUGCS,
\cite{2002ApJ...567..716R,2007A&A...466..805C}).\footnote{Note that
  we have added the Virgo cluster to the sample that we nevertheless
  refer to as extended HIFLUGCS catalogue in the following.}  For
previous work related to DM in clusters, see, e.g.,
\cite{2006A&A...455...21C,2009PhRvD..80b3005J,2011arXiv1104.3530S,2011ApJ...726L...6C,2010JCAP...05..025A}
and for recent work on CRs in clusters see
\cite{2010MNRAS.409..449P,2011MNRAS.410..127B,2008MNRAS.385.1211P,2009JCAP...08..002K,2010MNRAS.407.1565D}.

@@

In supernovae remnants and on scales of galaxies, there are convincing
evidences of non-thermal populations. Especially, in the MW, the
cosmic rays are observed directly as well as indirectly through radio,
X-ray, and gamma-ray emission. On larger scales of the order of few
100~kpc up to Mpc, there are currently a vast number of observations
of radio emission coming from radio mini halos in the center of
cooling flow clusters, radio relics in the periphery of clusters
\cite{2004rcfg.proc..335K} as well as smooth giant radio halos that
have been observed in more than 50 clusters
\cite{2003ASPC..301..143F,2008SSRv..134...93F}. This type of emission
is expected in clusters since the formation process of a galaxy
cluster is a very energetic processes that induces both turbulence as
well as frequently occurring merging and accretion shocks which both
are thought to accelerate relativistic non-thermal protons and radio
emitting electrons to high energies. Thus most of the observed radio
emission in clusters is expected to emerge from cosmic-ray electron
induced radio synchrotron emission. The precise origin of these
electrons, in especially relics and giant radio halos, is still not
settled. One possible scenario for the production of these electrons
are hadronic CR interactions with ambient gas protons which results in
the production of both charged and neutral pions that decay into
electrons, neutrinos, and gamma-rays (see
\cite{1980ApJ...239L..93D,1982AJ.....87.1266V,1999APh....12..169B,2000A&A...362..151D,2001ApJ...559...59M,2003MNRAS.342.1009M,2004A&A...413...17P,2004MNRAS.352...76P,2008MNRAS.385.1211P,2008MNRAS.385.1242P,2009JCAP...09..024K,2010MNRAS.401...47D,2010MNRAS.407.1565D,2010ApJ...722..737K,2011A&A...527A..99E}
).\footnote{An alternative scenario is the second order turbulent
  re-acceleration through the interaction of a previously injected
  relativistic population of electrons by supernova driven winds or
  active galactic nuclei with plasma waves and
  magneto-turbulence. \protect
  \cite{1987A&A...182...21S,1993ApJ...406..399G,2004MNRAS.350.1174B,2005MNRAS.363.1173B,2007MNRAS.378..245B,2011MNRAS.410..127B,2009A&A...507..661B}}
Supporting evidence comes from the smoothness of the extended radio
emission that often resembles that observed in thermal X-rays. This
can be easily achieve in the hadronic model since the long cooling
time of cosmic ray protons (CRs) allows for an extended and large
population of CRs to build up over the cluster history
\cite{1996SSRv...75..279V, 1997ApJ...477..560E,
  1997ApJ...487..529B}. The production of these secondaries depend
both on the gas and CR densities in the cluster, where the CR
density roughly traces the gas outside the core regime and is slightly
enhanced in the center.  This density scaling implies that clusters
are great targets for Cherenkov telescopes with a high sensitivity for
the central parts of nearby clusters. Detecting the cluster gamma-ray
emission is crucial in this respect as it potentially provides the
unique and unambiguous evidence of CR populations in clusters through
observing the $\pi^0$ bump at about 100~MeV in the spectra.

@@

As expected, the analytical CR model accounts for the ``scatter'' to
the scaling relation and biases the gamma-ray flux high for prominent
cool core clusters by up to a factor of 10 relative to the expectation
from the scaling relation. This is due to the high target gas
densities and associated CR densities in those cool cores that shape
the spatial emission characteristic to be very similar to that
observed in thermal X-rays. We caution that these predictions only
apply for the case of negligible active CR transport such as CR
streaming and diffusion relative to the gas. Gamma-ray fluxes might be
considerably smaller in those cluster that do not show an extended
diffuse radio(-mini) halo that might point to a centrally concentrated
CR distribution \cite{2011A&A...527A..99E}. The CR spectrum $E^2
dN/dE$ shows the characteristic pion bump at energies around
1~GeV. \C{It is expected to dominate the DM annihilation signal of BM
  models for clusters/groups, although we find that the gamma-ray flux
  induced by the BM is higher for about 1/5 of the clusters in the
  HIFLUGCS catalogue, where Fornax and M49 have the highest signal to
  noise. We find that \Fermi-LAT will be able to start constraining
  these DM models within five years using clusters of galaxies.}
\del{It is expected to dominate the DM annihilation signal of BM
  models for all clusters/groups but Fornax.}

Combining extended \Fermi upper limits and our model flux predictions,
we limit the relative CR pressure, $X_\CR = P_\CR/P_\rmn{th}$, in 32
nearby bright galaxy clusters of the \Fermi sample. \C{The best limits
  are found in Norma and Coma of the order of 3\%, with typical limits
  around 10\%. This is comparable to those inferred by the \Fermi
  collaboration \cite{2010ApJ...717L..71A} and mainly due to the
  differently assumed Bayesian prior on the CR pressure which implies
  a different cluster mass dependence of the resulting limits on
  $X_\CR$.}
